---
title: "Stat Interfence: Exp Distribution"
author: "Peter Sedivec"
date: "Saturday, February 21, 2015"
output: html_document
---

# Investigation of exponential distribution (rexp)

```{r echo=FALSE}
mexs = NULL
for (i in 1 : 1000) mexs = c(mexs, mean(rexp(40,0.2)))
```
If we compare the sample mean of mexs with the theoretical mean (1/lambda) we see that with 1000 samples the mean has converged very close to the theoretical mean:
```{r}
mean(mexs) # sample mean
1/0.2 # theoretical mean
mean(mexs) - 1/0.2
```
If we now look at the sample variance and compare it with the theoretical variance 1/lambda^2 we see that the sample variance is tighter (this doesn't really make sense)
```{r}
var(mexs) # sample variance
1/(0.2^2) # theoretical variance
```
If we look at a histogram of the distribution we can see that it is approximately normal
```{r}
hist(mexs, 40)
```
And as we increase the sample size from n=1,000 to n=10,000 we see the histogram converging to a normal distribution as expected per the central limit theorem
```{r echo=FALSE}
mexs2 = NULL
for (i in 1 : 10000) mexs2 = c(mexs2, mean(rexp(40,0.2)))
hist(mexs2, 40)
```